---
title: "Transcriptomics and the analysis of RNA-seq data"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Import today's data

The data for this hands-on session comes from a published RNA-seq experiment where airway smooth muscle cells were treated with dexamethasone, a synthetic glucocorticoid steroid with anti-inflammatory effects (Himes et al. 2014).

Glucocorticoids are used, for example, by people with asthma to reduce inflammation of the airways. The anti-inflammatory effects on airway smooth muscle (ASM) cells has been known for some time but the underlying molecular mechanisms are unclear.

```{r}
counts <- read.csv("airway_scaledcounts.csv", stringsAsFactors = FALSE)
metadata <-  read.csv("airway_metadata.csv", stringsAsFactors = FALSE)
```

How many genes are we dealing with?

```{r}
nrow(counts)
```

We want to calculate the average counts per gene for control samples and te for the drug treated samples. We will then compare theses values to see if there is a difference overall between control and treated samples for a given gene

This is our 'metadata' info that tells me where our control and treated samples are (i.e. which columns in 'counts')

```{r}
metadata
```

I want to access the control only rows in the metadata

```{r}
control <- metadata[metadata$dex == "control", ]
control
# metadata$id[metadata$dex == "control"]
```

Now use the $id of the control data to access the control columns in counts and calulate the average
```{r}
control.mean <- rowSums(counts[, control$id]) / nrow(control)
names(control.mean) <- counts$ensgene
```

Now I will do the same for treated data

```{r}
treated <- metadata[metadata$dex == "treated", ]
treated
```

```{r}
treated.mean <- rowSums(counts[, treated$id]) / nrow(treated)
```

We will combine our meancount data for bookeeping purposes

```{r}
meancounts <- data.frame(control.mean, treated.mean)
head(meancounts)
```

Let's plot control.mean vs treated.mean

```{r}
plot(meancounts)
```

Let's make this plot a log-log plot

```{r}
plot.default(meancounts, log = "xy")
```

We can find candidate differentially expressed genes by looking for genes with a large change between control and dex-treated samples. We usually look at the log2 of the fold change, because this has better mathematical properties.

Here we calculate log2foldchange, add it to our meancounts data.frame and inspect the results either with the head() or the View() function for example.

```{r}
meancounts$log2fc <- log2(meancounts[,"treated.mean"]/meancounts[,"control.mean"])
head(meancounts)
```

Remove any gene that has zero average counts

```{r}
x <- data.frame(c(1, 1, 10, 0), c(0, 1, 10, 0))
#x
#x == 0
ind <- unique(which(x == 0, arr.ind = TRUE)[, "row"])
x[ind, ]
```


```{r}
zero.vals <- which(meancounts[,1:2]==0, arr.ind=TRUE)

to.rm <- unique(zero.vals[,1])
mycounts <- meancounts[-to.rm,]
head(mycounts)
```

How many genes do I have left?

```{r}
nrow(mycounts)
```

A common threshold used for calling something differentially expressed is a log2(FoldChange) of greater than 2 or less than -2. Let’s filter the dataset both ways to see how many genes are up or down-regulated.

```{r}
up.ind <- mycounts$log2fc > 2
down.ind <- mycounts$log2fc < (-2)
```

```{r}
sum(up.ind)
sum(down.ind)
```

## Do this properly with DESeq2

```{r}
library(DESeq2)

```

Setup input for DESeq2 by specifying countdata, metadata and the design (i.e. which comparisons we are interested in)

```{r}
dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex, 
                              tidy=TRUE)
dds
```

```{r}
dds <- DESeq(dds)
```

```{r}
results(dds)
```

```{r}
res <- results(dds)
res
```

```{r}
plot(res$log2FoldChange, res$padj)
```

```{r}
plot(res$log2FoldChange, log(res$padj))

```

Let's plot this the way the rest of the world does

```{r}
plot(res$log2FoldChange, -log(res$padj),)

```

We need to add 'sig' column to our 'res' data. This will evaluate to TRUE if padj < 0.05 AND the absolute log2FoldChange >2, FALSE if not the case.

```{r}
res$sig <- res$padj < 0.05 & abs(res$log2FoldChange) > 2
```

If padj is also NA

```{r}
sum(is.na(res$sig))
```

```{r}
palette(c("gray", "blue"))
plot(res$log2FoldChange, -log(res$padj),
     col = res$sig+1)

```

```{r}
mycols <- rep("gray", nrow(res))
mycols[res$log2FoldChange > 2] <- "blue"
mycols[res$log2FoldChange < -2] <- "blue"
plot(res$log2FoldChange, -log(res$padj), col = mycols)
  abline(v = c(-2, 2), col = "darkgray", lty = 2)
  abline(h = -log(0.1), col = "darkgray", lty = 2)
```










